---
title: "Amount by Group function"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amount by Group function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
<<<<<<< HEAD

library(RDBESvisualise)
library(RDBEScore)
library(data.table)

=======
>>>>>>> Ana
```

## Introduction 

`amountByGroup()` is a utility function designed to summarise data from a single data table by one or more grouping variables, and it requires the user to specify the variables of interest to produce the outputs. 

It is particularly suited for workflows where:
  * Tables are previously prepared (e.g. joins, recoding, lookups done before)
  * Both numeric and categorical variables need to be summarised
  * The same summarisation logic should be reused across different tables
  
The function works on one table at a time and produces either summary tables
or simple bar plots.

It supports:

* Numeric variables -> summarised by `sum()`, with optional percentage conversion (`asPct = TRUE`).
* Categorical variables -> summarised as frequency counts.
* Multiple variables can be summarised at once:
 + Numeric variables -> aggregated together in *wide format*.
 + Categorical variables -> produce frequency tables by group.
* Filters can be applied prior to summarisation using a named list.
* Results can be optionally assigned to the global environment using `assign_to_global = TRUE`.
* Output options:
 + `"table"`: returns a data.table (or list of tables if mixed types). (Output type used by default)
 + `"plot"`: generates one or more ggplot2 visualisations.
* Safe use inside packages 

**Note**: The function intentionally does not perform joins or lookups — all preparation
(e.g. species code → species name) should be done before calling it.

<<<<<<< HEAD
=======
```{r}
library(RDBEScore)
library(RDBESvisualise)
```

>>>>>>> Ana

### Example data

For illustration, small examples of reproducible datasets, similar to RDBES data, are created.

```{r example data}
<<<<<<< HEAD
=======
library(data.table)
>>>>>>> Ana

CL <- data.table(
  Year = c('2021', '2021', '2021', '2021', '2022', '2022', '2021', '2021'),
  Quarter = c(1, 1, 2, 1, 2, 2, 1, 2),
  Country = c('ZW','ZW','ST','ST','ST','YM','YM','YM'),
  Species = c("COD", "HKE", "COD", "COD", "HKE", "HKE", "COD", "HKE"),
  WeightKg = c(100, 50, 200, 120, 80, 30, 10, 100),
  ValueEUR = c(300, 150, 600, 360, 240, 100, 50, 150)
  )

CE <- data.table(
  Year = c('2021', '2021', '2021', '2021', '2022', '2022', '2021', '2021'),
  Quarter = c(1, 1, 2, 1, 2, 2, 1, 2),
  Country = c('ZW','ZW','ST','ST','ST','YM','YM','YM'),
  Area = c('34.7','34.5','34.7','34.5','34.5','34.7','34.7','34.5'),
  NumDomTrip = c(3500, 300, 2010, 500, 250, 2600, 5000, 1000)
)

SA <- data.table(
  Year = c('2021', '2021', '2021', '2022', '2022'),
  Quarter = c(1, 1, 2, 1, 2),
  Species = c("COD", "HKE", "COD", "COD", "HKE"),
  CatchCategory = c("Lan", "Lan", "Dis", "Lan", "Dis"),
  totalWtMes = c(100, 50, 20, 120, 80),
  sampWtMes = c(20, 50, 20, 120, 30)
)
```


```{r print datasets}
print(CL)

print(CE)

print(SA)
```

### 1. Summarise a single numeric variable

The simplest use case is to sum a numeric column by one grouping variable.

```{r single numeric}

res_num <- amountByGroup(
data = CL,
var = "WeightKg",
valBy = "Year"
)

res_num

```

This returns a `data.table` with total weight per year.


### 2. Summarise a single categorical variable

If the variable is non-numeric, `amountByGroup()` automatically computes
frequencie counts.

```{r single categorical}

res_cat <- amountByGroup(
data = CL,
var = "Species",
valBy = "Year"
)

res_cat

```

Each row represents a combination of categorical variable (`Species`) and the grouping variables (`Year`), with a frequency column (`Freq`).

<<<<<<< HEAD
=======

>>>>>>> Ana
### 3. Multiple numeric variables (wide table)

Multiple numeric variables can be summarise at once, producing a *wide* table.

Example with CL table: 

```{r multiple numeric CL}

CLres_multi_num <- amountByGroup(
data = CL,
var = c("WeightKg", "ValueEUR"),
valBy = c("Year", "Species")
)

CLres_multi_num

```

Example with SA table:

```{r multiple numeric SA}

SAres_multi_num <- amountByGroup(
data = SA,
var = c("totalWtMes", "sampWtMes"),
valBy = c("Species", "CatchCategory")
)

SAres_multi_num

```

Each numeric variable is summed independently by the grouping variables.


### 4. Mixed numeric and categorical variables

Numeric and categorical variables can be supplied together.

```{r mixed variables}

res_mixed <- amountByGroup(
data = CE,
var = c("NumDomTrip", "Area"),
valBy = "Year"
)

str(res_mixed)

res_mixed

```


In this case, the function returns a list:

  * `$numeric` → summed numeric variables
<<<<<<< HEAD

=======
>>>>>>> Ana
  * `$categorical` → frequency tables for categorical variables

This keeps outputs explicit and avoids ambiguous table structures.


### 5. Applying filters

Filters are applied before summarising.

```{r applying filters}

res_filtered <- amountByGroup(
data = CE,
var = "NumDomTrip",
valBy = "Country",
filters = list(Year = 2021)
)

res_filtered

```

Filters use standard `%in%` logic and may contain multiple values: 

```{r multiple filters}

filters = list(Year = c(2021, 2022))

```

<<<<<<< HEAD
=======

>>>>>>> Ana
### 6. Percentages instead of absolute values

For numeric variables, results can be expressed as percentages of the total.

Example with SA table:
```{r percentages SA}

SA_res_pct <- amountByGroup(
data = SA,
var = "totalWtMes",
valBy = "CatchCategory",
asPct = TRUE
)

SA_res_pct

```

Example with CE table:

```{r percentages CE}

CE_res_pct <- amountByGroup(
data = CE,
var = "NumDomTrip",
valBy = "Area",
filters = list(Year=2021),
asPct = TRUE
)

CE_res_pct

```

Percentages are computed per variable, not per group.

### 7. Plot output

Setting `output_type = "plot"` produces bar charts.

Example with SA table: 

```{r plot numeric SA}

plot_SA <- amountByGroup(
data = SA,
var = "sampWtMes",
valBy = "Species",
output_type = "plot"
)

plot_SA

```

Example with CE table: 

```{r plot numeric CE}

plot_CE <- amountByGroup(
data = CE,
var = "NumDomTrip",
valBy = "Country",
output_type = "plot"
)

plot_CE

```

Notes:

<<<<<<< HEAD
  * Numeric variables → bar plot of summed values

  * Categorical variables → bar plot of frequencies

  * When multiple variables are supplied, one plot per variable is generated

=======
  * Numeric variables → bar plot of summed value
  * Categorical variables → bar plot of frequencies
  * When multiple variables are supplied, one plot per variable is generated
>>>>>>> Ana
  * The function still invisibly returns the underlying summary table


### 8. Plotting multiple variables

When `output_type = "plot"` and more than one variable is supplied in `var`,
`amountByGroup()` produces one plot per variable, printed sequentially.

This applies to:

  * Multiple numeric variables
<<<<<<< HEAD

=======
>>>>>>> Ana
  * Mixed numeric + categorical variables

Each plot is generated independently to keep the output clear and interpretable.

<<<<<<< HEAD
=======

>>>>>>> Ana
#### 8.1 Multiple numeric variables

When all variables in `var` are numeric, one bar plot is produced for each variable,
using the same grouping variables.

```{r plot multiple numeric}

plot_multi_numeric <- amountByGroup(
data = CL,
var = c("WeightKg", "ValueEUR"),
valBy = c("Year", "Country"),
output_type = "plot"
)

plot_multi_numeric

```

This results in:

  * One plot for `WeightKg` by `Year` (faceted by `Country`)
  * One plot for `ValueEUR`  by `Year` (faceted by `Country`)

The plots are printed in sequence and the underlying summary table is still returned invisibly.


#### 8.2 Mixed numeric and categorical variables

Numeric and categorical variables can be plotted together, and each variable is handled according to its type:
  * Numeric variables → summed bar plots
  * Categorical variables → frequency (count) bar plots
  
```{r plot mixed variables}

plot_mixed_var <- amountByGroup(
data = CL,
var = c("WeightKg", "Species"),
valBy = "Year",
output_type = "plot"
)

plot_mixed_var

```  
  
This produces:
  1. A bar plot of total `WeightKg` by `Year`
  2. A bar plot of counts of `Species` by `Year`

This explicit separation avoids mixing incompatible summaries in the same figure.  
  
##### Faceting behaviour

When more than one grouping variable is supplied in `valBy`:
  * The first grouping variable is used on the x-axis
  * Additional grouping variables may be used for faceting (when applicable)

For example:

```{r plot facet variables}

plot_facet <-  amountByGroup(
data = CE,
var = c("NumDomTrip"),
valBy = c("Country","Area"),
output_type = "plot"
)

plot_facet

```

This produces a plot where:
  * `xvar <- "Year"`
  * `facet_wrap(~ Quarter)` is now applied
  * Result: one panel per `Area` , as expected


### 9. Assigning results to the global environment

For interactive workflows, results can be automatically assigned:

```{r assign results}

amountByGroup(
data = CL,
var = "WeightKg",
valBy = "Year",
assign_to_global = TRUE
)

```

This creates an object named *tbl_WeightKg_by_Year*


### Design philosophy

  * One table in, one summary out
  * No hidden joins or recoding
  * Explicit handling of numeric vs categorical variables
  * Predictable output structures
  * Predictable and reusable outputs
  
The function is intended as a final summarisation layer, not as a data
preparation tool.


### When not to use this function

  * When you need joins or code-to-label translations
  * When you need model-based summaries
  * When you need long-format reshaping for downstream modeling

In those cases, prepare the data first, then call `amountByGroup()`.


<<<<<<< HEAD
=======

>>>>>>> Ana
#END
